Data Science Applied to Ag - Final Project - ML

### 2025-04-26 code/Draft_script_No_Cat_Features.qmd
d

@@

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

@@

```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
#library(caret)       # Other great library for Machine Learning 
```

@@

# Create recipe for data preprocessing
weather_recipe <- recipe(yield ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
  step_rm(
    year,       # Remove year identifier
    site,       # Remove site identifier
    matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
  ) %>%
  # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
  step_dummy(all_nominal_predictors()) %>%
  # Imputation for any numeric NAs   
  step_impute_median(all_numeric_predictors())

@@

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),  # Maximum depth of each tree
  min_n = tune(),  # Minimum samples required to split a node
  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
  set_engine("xgboost") %>% #specify engine 
  set_mode("regression")  # Set to mode
   # Total number of boosting iterations

        # Step size shrinkage for each boosting step

@@

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  size = 10
)

@@

@@

#parallel processing
registerDoParallel(cores = parallel::detectCores()-1)

@@

stopImplicitCluster()

@@

>>>>

Declaration of AI use: the following code chunk was inspired and subsequently modified on the basis of code initially generated by ChatGPT

@@

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "yield" in the group_by() to include it in the data frame because "yield" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

@@

```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
#install.packages("caret")

library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
library(caret)       # Other great library for Machine Learning 
```

@@

# Create recipe for data preprocessing
weather_recipe <- recipe(yield ~ ., data = weather_train) %>% 
  # Remove identifier columns and months not in growing season
  step_rm(
    year,       # Remove year identifier
    site,       # Remove site identifier
    hybrid,     # Remove site identifier
    matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
  ) %>%
  # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
  step_dummy(all_nominal_predictors())

@@

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune() # ,  # Maximum depth of each tree
  #min_n = tune(),  # Minimum samples required to split a node
  #learn_rate = tune()
  ) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
  set_engine("xgboost") %>% #specify engine 
  set_mode("regression")  # Set to mode

@@

xgb_grid <- grid_latin_hypercube(
  trees(),
  tree_depth(),
  #min_n(),
  #learn_rate(),
  size = 10
)

@@

Note: the following code chunk would not work if "tree_depth" and "learn_rate" were not fine tuned in the "xgb_spec" object.

@@

#parallel processing
registerDoParallel(cores = parallel::detectCores()-1) #starts parallel processing

@@

stopImplicitCluster() #ends parallel processing

@@

## 4. Select Best Models

We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:

```{r}
# Based on lowest RMSE
best_rmse <- xgb_res %>% 
  select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse

```{r}
# Based on lowers RMSE within 1% loss
best_rmse_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rmse",
                     limit = 1
                     )%>% 
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss
```

```{r}
# Based on lowest RMSE within 1 se
best_rmse_one_std_err <- xgb_res %>% 
  select_by_one_std_err(metric = "rmse",
                        eval_time = 100,
                        trees
                        )%>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```

Here we use all three methods which we learn in this class for R2.

```{r}
# Based on greatest R2
best_r2 <- xgb_res %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

```{r}
# Based on lowers R2 within 1% loss
best_r2_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rsq",
                     limit = 1
                     ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```

```{r}
# Based on lowest R2 within 1 se
best_r2_one_std_error <- xgb_res %>% 
  select_by_one_std_err(metric = "rsq",
                        eval_time = 100,
                        trees
                        ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```

## Compare and Finalize Model

```{r comparing values}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error)
```

## 5. Final Specification

```{r final_spec_fit}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth #, # Maximum depth of each tree
  #min_n = best_r2$min_n,           # Minimum number of samples to split a node
  #learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

## 6. Final Fit and Predictions

## Validation

```{r final_fit}
set.seed(10)
final_fit <- last_fit(final_spec,
                weather_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```

## 7. Evaluate on *Test Set*

```{r final_fit_metrics}
final_fit %>%
  collect_metrics()
```

## 8. Evaluate on Training Set

```{r}
final_spec %>%
  fit(yield ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rmse(yield, .pred) %>%
  bind_rows(

# R2
final_spec %>%
  fit(yield ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rsq(yield, .pred))
```

## 9. Predicted vs Observed Plot

```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous() +
  scale_y_continuous() 
```

## 10. Variable Importance

```{r final_spec}
final_spec %>%
  fit(yield ~ .,
         data = bake(weather_prep, weather_train)) %>% #There little change in variable improtance if you use full dataset
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)
```

# End of XGboost

# Start of Random Forest

```{r}
#| message: false
#| warning: false

#install.packages("ranger") #one of the packages in R that models random forest

library(tidymodels)
library(tidyverse) #takes a while to load
library(vip)
library(ranger)
library(finetune)
```

## 2. Random Forest Training

### a. Random Forest Model specification

First, let's specify:  
  - the **type of model** we want to train  
  - which **engine** we want to use  
  - which **mode** we want to use

Random forest **hyperparameters**:  
  - **trees**: number of trees in the forest    
  - **mtry**: number of random features [or columns] sampled at each node split    
    - **min_n**: minimum number of data points in a node that are required for the node to be split further

[Important note: we are not gonna fine tune "min_n" in this exercise because of time constraint, but we should fine tune "min_n" to do random forest in our own data]

Let's create a model specification that will **fine-tune** the first two for us.

A given model type can be fit with different engines (e.g., through different packages). Here, we'll use the **ranger** engine/package.

```{r rf_spec}

rf_spec <- 
  # Specifying random forest as our model type, asking to tune the hyperparameters
  rand_forest(trees = tune(),
              mtry = tune() #both "trees" and "mtry" are model-type level hyperparameters, meaning that we fine tune "trees" and "mtry" inside "rand_forest()"
              ) %>%
    # Specify the engine (= package)
    set_engine("ranger") %>% #specifying "ranger" as the engine/package to run random forest 
    # Specifying mode  
    set_mode("regression") #random forest can handle both regression (when y is numerical) and classification (when y is categorical) #Here, we are specifying "set_mode("regression")" because our y variable is numerical [continuous]

rf_spec

### b. Hyper-parameter tuning

For our iterative search, we need:  
  - Our model specification (`rf_spec`)  
  - The recipe (`weather_recipe`)  
  - Our **resampling strategy** (don't have yet)

Let's define our resampling strategy below, using a 5-fold cross validation approach:

```{r resampling_foldcv}

set.seed(34549)

rf_resampling_foldcv <- vfold_cv(weather_train, 
                              v = 5) #but at least 10 folds are recommended

rf_resampling_foldcv
rf_resampling_foldcv$splits[[1]]
rf_resampling_foldcv$splits[[2]]

We will use an iterative search algorithm called **simulated annealing**.

In the algorithm below, we are asking for 10 iterations.

```{r rf_grid_result}

rf_grid_result <- tune_sim_anneal(object = rf_spec,
                     preprocessor = weather_recipe,
                     resamples = rf_resampling_foldcv,
                    #param_info = rf_param,
                    iter = 10 
                     )

rf_grid_result
rf_grid_result$.metrics[[2]]
```

Let's collect a summary of metrics (across all folds, for each iteration), and plot them.

First, RMSE (lower is better):

```{r RMSE}
rf_grid_result %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = mtry, 
             y = trees 
             )) +
  geom_path(group = 1) +
  geom_point(aes(color = mean),
             size = 3) + 
  scale_color_viridis_b() +
  geom_text(aes(label = .iter), nudge_x = .0005) +
  labs(title = "RMSE")
```

Now, let's look into R2 (higher is better):

```{r R2}

rf_grid_result %>%
  collect_metrics() %>%
  filter(.metric == "rsq") %>%
  ggplot(aes(x = mtry, 
             y = trees 
             )) +
  geom_path(group = 1) +
  geom_point(aes(color = mean),
             size = 3) + 
  scale_color_viridis_b() +
  geom_text(aes(label = .iter), nudge_x = .0005) +
  labs(title = "R2")

We want to see higher R2 (R2: higher is better).

The following code chunk will select the best model based on "percent loss" of the metric.

# Based on lowest RMSE
best_rmse <- rf_grid_result %>%
  select_by_pct_loss("trees",
                     metric = "rmse",
                     limit = 2 #"limit = 2": this is how many percent losses you are accepting, 2% in this case
                     )

best_rmse

The following code chunk will select the best model based on greatest "R2"

# Based on greatest R2
best_r2 <- rf_grid_result %>%
  select_by_pct_loss("trees",
                     metric = "rsq",
                     limit = 2
                     )

best_r2

I got the he following results that were different than Dr. Bastos's results (because of the randomness of 2 hyperparameters going on at each tree for a lot of models):

Based on RMSE, we would choose   
  - mtry = 25 
  - trees = 443

Based on R2, we would choose   
  - mtry = 22
  - trees = 113

Let's use the hyperparameter values that optimized R2 to fit our final model.

```{r final_spec}
final_spec <- rand_forest(trees = best_r2$trees,
                          mtry = best_r2$mtry) %>%
  # Specify the engine
  set_engine("ranger",
             importance = "permutation" #"permutation" is how we look into variable importance in random forest
             ) %>%
    # Specifying mode  
  set_mode("regression")

final_spec
```

## 3. Validation

Now that we determined our best model, let's do our **last fit**.

### What does last fit mean? (Will come in exam)

This means 2 things:  
  - Training the optimum hyperparameter values on the **entire training set**  
  - Using it to **predict** on the **test set**

These 2 steps can be completed in one function, as below:

```{r final_fit}
final_fit <- last_fit(final_spec, 
                weather_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```

Metrics on the **test set**:

```{r}
final_fit %>%
  collect_metrics()
```

From the output,
Fit metrics on the test set:
rmse = 2.9317445	
rsq	= 0.1882245

Metrics on **train set** (for curiosity and compare to test set):  
```{r}
# RMSE
final_spec %>%
  fit(yield ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rmse(yield, .pred) %>%
  bind_rows(

    # R2
    final_spec %>%
      fit(yield ~ .,
          data = bake(weather_prep, 
                      weather_train)) %>%
      augment(new_data = bake(weather_prep, 
                              weather_train)) %>% 
      rsq(yield, .pred)

  )

From the output,
Fit metrics on the train set:
rmse = 1.1385078	
rsq	= 0.9470296

How does metrics on test compare to metrics on train?

Fit metrics on the test set:
rmse = 2.9317445	
rsq	= 0.1882245

Fit metrics on the train set:
rmse = 1.1385078	
rsq	= 0.9470296

We see that the R2 is really r-e-a-l-l-y high in the train set (rsq	= 0.9470296), as compared to the R2 of the test set 0.1882245. The RMSE is lower in the train set (rmse = 1.1385078) as compared to the RMSE of the test set (rmse = 2.9317445). [This is the reason why it's okay to report the fit metrics of the train set to understand how well our model is explaining our data, but it is not a good measure of the predictive ability of the model. To understand the predictive ability of the model, we have to look at the fit metrics of the test set. Notice that the R2 of the test set was 0.1882245, whereas the R2 of the train set was 0.9470296 -- which is a very large difference. Hence, *we need to always report the fit metrics of the test set* along with the fit metrics of the train set. ]

Predicted vs. observed plot:  
```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(20, 40)) +
  scale_y_continuous(limits = c(20, 40)) 
```

[**Very important note:**
To interpret random forest model, we should always expcet to see (1) *perdicted vs. observed plot* and (3) *variable importance plot*, at the least -- these are the 2 things that must be mentioned in a random forest model. 
]

*Variable importance:*

[For Random Forest,] The importance metric we are evaluating here is **permutation**.

In the permutation-based approach, for each tree, the out-of-bag sample is passed down the tree and the prediction accuracy is recorded. [as is]

Then the values for each variable (one at a time) are randomly permuted and the accuracy is again computed.

The decrease in accuracy as a result of this randomly shuffling of feature values is averaged over all the trees for each predictor.

The variables with the **largest average decrease in accuracy** [after being permuted] are [the ones that are] considered **most important**.

final_spec %>%
  fit(yield ~ .,
         data = bake(weather_prep, weather)) %>%
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)

#End of Random Forest

#Start of getting yield prediction (yield_mg_ha) for the unseen data in "testing_submission.csv" file
